### 2022-11-21 introduction/introduction.tex
Added paragraph of protein structure and added Punnett square example

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify their details into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, a punnettSince life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

@@

Despite performing this diverse range of functions, all proteins are made up of a set of twenty simple building blocks called amino acids. Each amino acid in a protein is bonded to the next, forming a linear chain like beads on a string.

>>>>

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules. However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

@@

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids. Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination.

FOOTNOTE: The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the twenty "canonical" amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.
